---
title: "Data Transformation with 'dplyr'"
output: github_document
---

*Aim: Get familiar with using the [`dplyr`](https://dplyr.tidyverse.org/) package to wrangle dataframes.*


*Please see* ***[Data Transformation with dplyr cheat sheets](https://rstudio.com/resources/cheatsheets/)*** *for a nice overview of functions in this package.*

________________________________________________

Load in packages:

```{r, message = F, warning=FALSE}
# install.packages("dplyr")
# install.packages("tibble")
library(dplyr)
library(tibble)
```

In order to use `dplyr` functions you need to ensure your dataframe is in "tidy" format. This means in your dataframe each row represents an observation and each column represents a variable. This is also called "long format".

We will use the `mtcars` (Motor Trend magazine car road tests) dataset which is in tidy format to understand dplyr functions.
```{r}
head(mtcars)
```

Tidyverse includes an enhancement to `data.frame`s called a "tibble" that are useful for interactive data analysis. Its most useful feature is that when entered in the command prompt on its own, a `tibble` will by default print out only its first 10 rows and only as many column as can be displayed on a single line. Use `as_tibble` to convert a normal `data.frame` to a `tibble`

```{r}
mtcars <- as_tibble(datasets::mtcars)
mtcars
```



### Summarising data

<span style="color: darkblue;"> 
**How do I get summary of data?**
</span>

```{r}
summary(mtcars)

# Get mean of mpg
summarise(mtcars, avg=mean(mpg))

# What's the best way to get the mean for all columns?
```

<span style="color: darkblue;"> 
**How do I get summaries of grouped variables?**
</span>

```{r}
grouped_mtcars <- mtcars %>%
                    group_by(cyl) %>%
                    summarise(avg=mean(mpg))
grouped_mtcars
```

Pipe operator `%>%` can be used to write tidy code. You can see that the above, with pipes, is easier to read than the following, without pipes:
```{r}
summarise(group_by(mtcars, cyl), avg = mean(mpg))
```


### Manipulating cases ie. rows 

<span style="color: darkblue;"> 
**How do I subset rows based on a logical criteria?**
</span>
```{r}
filter(mtcars, mpg > 25)
```

<span style="color: darkblue;"> 
**How do I remove rows with duplicate values?**
</span>

```{r}
# Return all unique items for "cyl" column
distinct(mtcars, cyl)

# How do you keep the cases from the other columns?

# How can you ensure you're removing the correct duplicate cases? Presumbly it keeps the first unique case?
```

<span style="color: darkblue;"> 
**How do I select rows by position?**
</span>
```{r}
mtcars %>%
  slice(10:15)

# last 5 rows:
mtcars %>% 
  slice((nrow(mtcars)-5):nrow(mtcars))

# Is there a neater way of doing that?
```

<span style="color: darkblue;"> 
**How do I select top n entries of a variable and order them?**
</span>

```{r}
top_n(mtcars, 5, mpg) %>%
  arrange(desc(mpg))
```


### Manipulating variables ie. columns

<span style="color: darkblue;"> 
**How do I subset a dataframe by columns?**
</span>
```{r}
select(mtcars, mpg, cyl)

# To extract a column as a vector:
pull(mtcars, mpg)
```

<span style="color: darkblue;"> 
**Which function should I use if I want to add a column to my dataframe?**
</span>

```{r}
mutate(mtcars, gpm = 1/mpg)

# Can input any vector, as long as it's the same length as number of rows in your data frame. eg:
a_vector <- rep(1, nrow(mtcars))
mutate(mtcars, random_vec = a_vector)
```

<span style="color: darkblue;"> 
**How do I create a new column but drop the rest?**
</span>

```{r}
transmute(mtcars, gpm = 1/mpg)
```

<span style="color: darkblue;"> 
**How do I apply a function to every column?**
</span>

The function to apply can either be supplied directly (`fun`) or as an anonymous ("lambda") function using the formula syntax (`~fun(.)`)

```{r}
mutate_all(mtcars, ~.*100) # note the use of . in the lambda function

# What if not all our columns were numeric? We can do this:
mutate_if(mtcars, is.numeric, ~.*100)

# What if you want to log transform a few variables for example.
# eg. log transform the mpg and cyl columns only:
mutate_at(mtcars, vars(c(mpg, cyl)) , log)

```


<span style="color: darkblue;"> 
**How do I rename columns?**
</span>

```{r}
rename(mtcars, miles_per_gallon=mpg)

# Be careful about names, avoid spaces. If you have have spaces use `` to enclose the variable name.
rename(mtcars, `miles per gallon`=mpg) 
```

### Row names
Be aware that if you have a variable stored as a row name then that you want to work with then you need to move it to a column. The `tibble` package has a function for this.
```{r}
rownames_to_column(datasets::mtcars, var="Car") 
```
We can also do this conversion when making the `tibble`
```{r}
as_tibble(datasets::mtcars, rownames='Car')
```

### Combining Tables

`dplyr` also has functions for combining data frames:

```{r}
# New dataframe:
dataf <- select(mtcars, new_mpg=mpg)
bind_cols(mtcars, dataf) # BE SURE ROWS ALLIGN
```
May be better to use a "mutating join" as that will match values with the rows they correspond to, ie. if your rows don't allign.
These include: `left_join`,`right_join`, `inner_join` and `full_join`. 
Use the `by=` argument to specify the column(s) both data frames match on or a named vector if the columns have different names in each data frame.

Similarly, there are functions for rows. Such as `bind_rows`, `intersect`, `setdiff`, `union`.


# Advantages and disadvantages of dplyr (and the "Tidy") system

Pros:

- Consistent grammar for extracting, filtering, viewing, and manipulating data tables.
- [Verb-oriented data manipulation](https://dplyr.tidyverse.org/articles/base.html) what each processing step is aimed at doing
- "Piping" approach makes it easier to build up operations one step at a time
- Encourages copying data to a new data.frame when using mutations rather than modifying data in place, which can introduce errors in interactive analysis when lines of code are run out-of-order.

Cons:

- Function names, syntax, and defaults seem to change every few years, breaking old code.
- Namespace conflicts with several standard R functions (`stats::filter`, `stats::lag`, `MASS::select`, `base::intersect`, `base::union`)
- Extra syntax required for indirection, when you want to do something with a column whose name is stored in a variable rather than having to name the column directly. This happens for example when your script needs to select a column or create a new column but you don't know what the column is called until the script is run. 



