---
title: "String manipulation and regular expressions"
output: github_document
---

*Aim: Learn how to extract information from strings (text and character data) using regular expressions*

We will look at some built-in fucntions for working with strings (`substr`, `strsplit`, `paste`) as well as their Tidy equivalents from the [`tidyr`](https://tidyr.tidyverse.org) and [`stringr`](https://stringr.tidyverse.org) packages. See also: [R For Data Science Chapter 14: Strings](https://r4ds.had.co.nz/strings.html).

# String manipulation

A "string" is a sequence of symbols. In computer languages, strings are typically a data structure for storing text information as a sequence of letters, numbers, punctuation, and other language symbols. In R and most programming languages, strings are demarcated by surrounding them by single `'` or double `"` quotation marks.

```{r}
"I like cats."
"猫が大好き。"
```

In R, single or double quotation marks are functionally equivalent but they have different meanings in other languages. For example in the C programming language, single quotes are used to demarcate single characters while double quotes must be used for strings. In bash, a string in single quotes will be interpreted literally (as-is):

```{bash}
echo 'Hi, $USER, the date is $(date +%Y-%m-%d)'
```
while a string in double quotes allows string interpolation (where bash will look through the string for meaningful punctuation and interpret it as a command)
```{bash}
echo "Hi, $USER, the date is $(date +%Y-%m-%d)"
```

In R, strings are stored as type `character`
```{r}
typeof("Hello, World")
```

Outside of string processing and text data storage, the main place in R that `character` data shows up is in indexing into named elements of vectors and lists (including `data.frame`s). E.g., a named vector:

```{r}
sample_sizes <- c(cases=11795, controls=14084)
```

The elements can either be indexed with string _literals_
```{r}
sample_sizes[["cases"]]
sample_sizes[["controls"]]
```

or with a variable assigned the value we want to extract
```{r}
a <- "cases"
sample_sizes[[a]]
```

## String escapes

## Built-in commands

Three common operations that it is worthwhile knowing the built-in commands for are counting, subsetting, splitting, and combining strings.

### Getting information about strings (`nchar`)

The number of characters in a string can be calculated with `nchar`

```{r}
nchar("hello")
```
This function can also processing vectors of characters
```{r}
nchar(c("hello", "abracadabra"))
```

**Quick question**: What information is `length("hello")` returning?

### Subsetting strings (`substr`)

Parts of a string can be extracted by character position using `subtr` (substring). The first two characters of the word "hello" can be extracted with
```{r}
substr("hello", start=3, stop=5)
```
Getting a part of a string using position is useful if all of the strings you want to work with are of uniform length and formatting. For example if you have a list of ICD-10 diagnoses:

```{r}
icd10 <- c("F410", "F322", "I119", "E112")
```
and you want to extract information about which chapter the diagnoses are from
```{r}
table(substr(icd10, 1, 1))
```

`substr` can also be used for assignment
```{r}
classic_pub <- "The Sheep Head Inn"
substr(classic_pub, 13, 13) <- 'i'
classic_pub
```

### Cleaning up whitespace (`trimws`)

### String substitution (`gsub`)

### String transformation (`toupper` and `tolower`)

### String separation (`strsplit`)

### String concatenation (`paste`)

## Tidyverse: `stringr`

# Regular expressions

Given a particular string...