---
title: "String manipulation and date parsing"
output: github_document
---

*Aim: Learn how to process strings (text and character data) and parse dates*

We will look at some built-in fucntions for working with strings (`substr`, `strsplit`, `paste`) as well as their Tidy equivalents from the [`tidyr`](https://tidyr.tidyverse.org) and [`stringr`](https://stringr.tidyverse.org) packages. See also: [R For Data Science Chapter 14: Strings](https://r4ds.had.co.nz/strings.html).

# String manipulation

A "string" is a sequence of symbols. In computer languages, strings are typically a data structure for storing text information as a sequence of letters, numbers, punctuation, and other language symbols. In R and most programming languages, strings are demarcated by surrounding them by single `'` or double `"` quotation marks.

```{r}
"I like cats."
"猫が大好き。"
```

In R, single or double quotation marks are functionally equivalent but they have different meanings in other languages. For example in the C programming language, single quotes are used to demarcate single characters while double quotes must be used for strings. In bash, a string in single quotes will be interpreted literally (as-is):

```{bash}
echo 'Hi, $USER, the date is $(date +%Y-%m-%d)'
```
while a string in double quotes allows string interpolation (where bash will look through the string for meaningful punctuation and interpret it as a command)
```{bash}
echo "Hi, $USER, the date is $(date +%Y-%m-%d)"
```

In R, strings are stored as type `character`
```{r}
typeof("Hello, World")
```

Outside of string processing and text data storage, the main place in R that `character` data shows up is in indexing into named elements of vectors and lists (including `data.frame`s). E.g., a named vector:

```{r}
sample_sizes <- c(cases=11795, controls=14084)
```

The elements can either be indexed with string _literals_
```{r}
sample_sizes[["cases"]]
sample_sizes[["controls"]]
```

or with a variable assigned the value we want to extract
```{r}
a <- "cases"
sample_sizes[[a]]
```

## String escapes

The majority of alphabetic, numeric, and punctuation characters are represented in a string just as they appear. Strings can also contain special characters take on a meaning other than a literal character. These are called "escape characters" (or, if they involve more than one character, "escape sequences") and are prepended with `\` (backslash). The most common escape characters that you'll encounter in data processing are:

- `\"` and `\'` to specify a double or single quote within a string.
```{r}
"This double-quoted string needs to escape \"double\" quotes"
'This single-quoted string needs to escape \'single\' quotes'
```
- `\n` for line breaks (newline)
```{r}
string_with_newlines <- "This\nstring\nhas\nseveral\nlines."
```
- `\t` for tabs
```{r}
string_with_tabs <- "This\tstring\tis\ttab\tseparated"
```
- `\u` and `\U` for Unicode characters.
```{r}
string_with_unicode <- "\U0001f978"
```


# Displaying strings

When echoed in the terminal, strings in R will be output just as they appear in the source code, exactly as they were typed, including the escape characters. To print an interpreted version of the string to the screen, use the `cat()` function:

**Newlines**
```{r}
string_with_newlines
cat(string_with_newlines)
```

**Tabs**
```{r}
string_with_tabs
cat(string_with_tabs)
```

**Unicode**
```{r}
string_with_unicode
cat(string_with_unicode)
"💀👻🐶🐱"
cat("💀👻🐶🐱")
```

# Built-in commands

Four common operations that it is worthwhile knowing the built-in commands for are counting, subsetting, splitting, and combining strings.

## Getting information about strings (`nchar`)

The number of characters in a string can be calculated with `nchar`

```{r}
nchar("hello")
```
This function can also processing vectors of characters
```{r}
nchar(c("hello", "abracadabra"))
```

**Quick question**: What information is `length("hello")` returning?

## Subsetting strings (`substr`)

Parts of a string can be extracted by character position using `subtr` (substring). The third, fourth, and fifth characters of the word "hello" can be extracted with
```{r}
substr("hello", start=3, stop=5)
```
Getting a part of a string using position is useful if all of the strings you want to work with are of uniform length and formatting. For example if you have a list of ICD-10 diagnoses:

```{r}
icd10 <- c("F410", "F322", "I119", "E112")
```
and you want to extract information about which chapter the diagnoses are from
```{r}
table(substr(icd10, 1, 1))
```

`substr` can also be used for assignment
```{r}
classic_pub <- "The Sheep Head Inn"
substr(classic_pub, 13, 13) <- 'i'
classic_pub
```

## Cleaning up whitespace (`trimws`)

Extraneous whitespace is usually not a problem in well-curated datasets, but it can crop up in older and ad-hoc files, particularly when data has been copied and pasted into a spreadsheet. `trimws()` can be used to _trim_ whitespace at the start or end of a string

```{r}
icd10_with_ws <- c("F41 ", " F32", "I11  ", "   E112")
trimws(icd10_with_ws)
```

## Finding strings (`grep`)

Strings can be searched with `grep()` (_g_lobally search for a _r_egular _e_xpression and _p_rint matching lines). For now we will just search for string literals. In a future session, we will look at more flexible searching using regular expressions. 

Unlike most commands, the object that `grep()` operates on (a character vector) is the second argument. The first argument is the thing being search for. For example, to find diagnoses inthe `icd10` vector that start with "F":
```{r}
grep(pattern="F", x=icd10)
```
Grep returns the indices of the elements that match, so it can be used to subset the array being searched through
```{r}
icd10[grep("F", icd10)]
```
A shortcut for doing the same thing is with the `value=TRUE` argument
```{r}
grep("F", icd10, value=TRUE)
```

`grepl()` returns `TRUE` or `FALSE` for each match

```{r}
grepl("F", icd10)
```
`grepl()` is therefore useful for things like `ifelse()` statements
```{r}
ifelse(grepl("F", icd10), yes='mental health', no='physical health')
```

## String substitution (`gsub`)

Finding and replace can be done with `sub()` (which will replace the first match) and `gsub()` (which will replace all matches).

```{r}
sub(pattern="Head", replacement="Heid", x=classic_pub)
```

## String transformation (`toupper` and `tolower`)

Capitals can be converted to lowercase and vice-versa:

```{r}
tolower("BE QUIET!!")
toupper("evertything looks like shouting when it's in allcaps")
```

String transformation and substitution can be useful for turning plain labels into variable names or filenames. For example, a list of names:

```{r}
buildings <- c("Ashworth Laboratories", "7 George Square", "Kennedy Tower")
building_labels <- tolower(gsub(" ", "_", buildings))
building_labels
```

For a more advanced version of this, see the [snakecase](https://cran.r-project.org/web/packages/snakecase/vignettes/caseconverters.html) package.

## String separation (`strsplit`)

A string can be divided into parts based on a specified `split` sequence. For example, splitting a sentence into separate words. 

```{r}
strsplit("The quick brown fox jumps over the lazy dog", split=" ")
```

It can be useful for extracting parts of strings that are formatted in a uniform way.
```{r}
sessions <- c('2020-09-30', '2020-10-23', '2020-10-28', '2020-11-11', '2020-12-09', '2021-01-27', '2021-02-24', '2021-03-10')
sessions_ymd <- strsplit(sessions, split='-')
sessions_ymd
```

`strsplit` returns a list of character vectors. 

## String concatenation (`paste`)

Strings can be combined together using paste.
```{r}
paste('This', 'is', 'a', 'sentence.')
```
By default the combined strings will be separated by a space but this can be changed with the `sep` argument.

```{r}
paste('2020', '09', '30', sep='-')
```

The arguments can also be vectors to produce multiple output strings

```{r}
paste(c('2020', '2021'), c('09', '02'), c('30', '24'), sep='-')
```
Note how the first elements of each input get combined together, then the second element of each input. If the inputs are of different lengths, the shorter ones will be recycled. This is useful for example when you need to add a prefix to a vector of strings:
```{r}
paste('chr', 1:22, sep='')
```
The vector of strings output by `paste` can themselves be concatenated together using the `collapse` argument
```{r}
paste(c('2020', '2021'), c('09', '02'), c('30', '24'), sep='-', collapse=', ')
```

A useful shortcut for `paste(..., sep='')` in the common instance where you don't want the elements to be separated by a character is `paste0(...)`
```{r}
paste0('chr', 1:22)
```

# Tidyverse: `stringr` and `tidyr`

The built-in string manipulation functions can be a bit clunky to work with because they don't all follow the same argument ordering conventions. Some functions are of the form `f(string, pattern)` while others are of the form `f(pattern, string)`. The [`stringr`](https://stringr.tidyverse.org) package is a great alternative because all the functions follow the format `f(string, pattern)`

```{r}
library(stringr)
```

Built-in                            | `stringr`
----------------------------------- | ------------------------------
`substr(string, start, stop)`       | `str_sub(string, start, stop)`
`trimws(str, which='both')`         | `str_trim(str)`
`grep(pattern, string)`             | `str_which(string, pattern)`
`grepl(pattern, string)`            | `str_detect(string, pattern)`
`grep(pattern, string, value=TRUE)` | `str_match(string, pattern)`
`sub(pattern, replacement, string)` | `str_replace(string, pattern, replacement)`
`toupper(string)`                   | `str_to_upper(string)`
`tolower(string)`                   | `str_to_lower(string)`
`strsplit(string, split)`           | `str_split(string, pattern)`
`paste(string, sep, collapse)`      | `str_c(string, sep, collapse)`

**Exercise**: revise these built-in commands with their `stringr` equivalents
```r
substr("hello", start=3, stop=5)
trimws(icd10_with_ws) 
grep("F", icd10, value=TRUE)
sub(pattern="Head", replacement="Heid", x=classic_pub)
strsplit("The quick brown fox jumps over the lazy dog", split=" ")
paste('2020', '09', '30', sep='-')
```

## String interpolation: `str_glue`

Another feature that comes with the `stringr` package is interpolation, where part of a string is substituted with the value from a variable. The variable is specified by surrounding in curly braces `{varname}`:

```{r}
session_name = 'Strings and Dates'
str_glue("This coding club is the {session_name} session.")
```

`str_glue_data` can be used to create interpolated strings where the variable values are pipped from a `data.frame` or `tibble`
```{r}
mtcars_tb <- tibble::as_tibble(mtcars, rownames='car')
mtcars_tb %>% str_glue_data("The {car} gets {mpg} miles per gallon.")
```


# Working with dates (`lubridate`)

One of the most common types of string data to encounter in research are dates. When dates are formatted in a uniform way, some information can be extracted using the string tools we've alredy seen:

```{r}
session_years <- as.numeric(sapply(str_split(sessions, pattern='-'), dplyr::first))
session_years
```

Dates are also important as the basis of calculations. For example, we might be interested in knowing how much time has passed between a baseline and repeat assessment, or we might want to group dates together to define hospitalisation or prescribing episodes. Date information in R can be stored in `Date` objects. The `lubridate` package adds additional functionality for parsing character data into `Date`s and doing calculations that properly handle parculiarities like leap years and time zones.

## Creating date objects and parsing dates

Under the the hood, dates are stored as number of days since January 1st, 1970 (the "[Unix epoch]"(https://en.wikipedia.org/wiki/Unix_time)). An object holding date/time information can be created with `ISOdate()``

```{r}
ISOdate(year=2021, month=3, day=10)
```

A basic date object can also be automatically parsed from a string that is formatted as `YYYY/MM/DD` or `YYYY-MM-DD`

```{r}
as.Date("2021/03/10")
as.Date("2021-03-10")
```

Lubridate has a similar function `ymd()` that is less picky about punctuation separating the different parts of the date string, as long as they appear in the order year–month–day.

```{r}
library(lubridate)
ymd("2021/03/10")
ymd("2021-03-10")
ymd("2021 03 10")
ymd("20210310")
```

Lubridate has other convenience functions if the date information is in a different order
```{r}
dmy("10/03/2021")
mdy("03/10/2021")
```

If the date information is not as nicely structured, it can be extracted using  `parse_date_time()`. This function's second argument is an `orders` vector of string encoding the different ordering and formatting of date information to look for. The orders string uses single or dual characters to encode the formatting of the date information:

- `a` Abbreviated weekday name in the current locale. (Also matches full name)
- `A` Full weekday name in the current locale.  (Also matches abbreviated name).
- `b` Abbreviated or full month name in the current locale.
- `d` Day of the month as decimal number (01-31 or 0-31)
- `m` Month as decimal number (01-12 or 1-12). 
- `w` Weekday as decimal number (0-6, Sunday is 0).
- `y` Year without century (00-99 or 0-99).
- `Y` Year with century.
- `Om` Matches numeric month and English alphabetic months (Both, long and abbreviated forms).

```{r}
parse_date_time("2021-03-10", orders="Ymd")
parse_date_time("Mar 10th 2021", orders="bmY")
parse_date_time("10 March 21", orders="dby")
```

If a vector of parsing orders are passed, the function will try each of the options
```{r}
parse_date_time(c("2021-03-10","Mar 10th 2021", "10 March 21", "Jan 1st 1970"), orders=c("bmY", "dby", "Ymd"))
```

## Calculations with dates

Once you have two dates, you can use subtraction to get the difference between them.

```{r}
unix_epoch = ymd("1970-01-01")
string_session = ymd("2021-03-10")
string_session - unix_epoch
```

Two dates can also define an **interval**
```{r}
epoch_session <- interval(unix_epoch, string_session)
epoch_session
```

An interval can be divided by a duration to convert them to a different unit. How many minutes between today and the Unix epoch?
```{r}
epoch_session / dminutes(1)
```
Or years?
```{r}
epoch_session / dyears(1)
```

**NB**: Lubridate has two ways of interpreting the length of an `interval()`: a _duration_ repesents absolute differences in time (a day is always 24 hours, a year is always 365 days) while a _period_ represents time in a way that respects how calendars account for them (some years are 366 days, etc). For _period_ the functions are plural time units while the equivalent _duration_ functions are prefixed with `d`

```{r}
string_session + months(1)
string_session + dmonths(1)
``` 