---
title: "Using the pipe (%>%) operator in R"
author: "Amelia Edmondson-Stait"
date: "25/11/2020"
output: rmarkdown::github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### What are pipes?
Only recently discovered how amazingly useful pipes are and finally got round to figuring out how to start using them and improve my code to become more efficient and readable.

**"The pipe operator will forward a value, or the result of an expression, into the next function call/expression."**

The `magrittr` package contains this useful pipe operator `%>%` discussed in this doc. 
Load package and create dummy GWAS summary statistic data to use in this doc.
```{r, echo = TRUE, error=FALSE, warning=FALSE, message=FALSE}
# Load libraries
library(tidyverse) # For %>%, also loads other useful packages (see resources links at end of this doc).
library(stringi) # For creating random character vectors
```

```{r}
# Create dummy data
CHR <- unlist(lapply(1:22, function(x) rep(x, 10)))
SNP <- paste("rs", round(runif(220, min=11250701, max=185339560),0), sep = "")
BP <- round(runif(220, min=151476, max=2622752),0)
A1 <- stri_rand_strings(220, 1, pattern = "[ATCG]")
A2 <- rep(NA, 220)
A2[A1 == "A"] <- "T"
A2[A1 == "T"] <- "A"
A2[A1 == "G"] <- "C"
A2[A1 == "C"] <- "G"
OR <- runif(220, min=0, max=1)
P <- runif(220, min=0.000000001, max=0.98347)

# Create dataframe
sumstats <- data.frame("CHR" = CHR, "SNP" = SNP, "BP" = BP, "A1" = A1, "A2" = A2, "OR" = OR, "P" = P)

head(sumstats)

```

### How can we use pipes?

In R, there are many ways of performing the same task. First I'll demonstrate a long-winded way without using pipes.
I would like to subset my dataframe to only contain chromosome 1 and sort this new dataframe based on ascending P  values. Sorting a dataframe based on the order of another column or vector is something I do alot.

```{r}
sumstatrs_chr1 <- sumstats[sumstats$CHR == 1,][order(sumstats[sumstats$CHR == 1,]$P), ]
sumstatrs_chr1
```

You can see that this did what I wanted it to, but it's really not an easy to read piece of code.
I could overcome this by creating intermediate variables. Such as:

```{r}
sumstatrs_chr1_int <- sumstats[sumstats$CHR == 1,]
sumstatrs_chr1 <- sumstatrs_chr1_int[order(sumstatrs_chr1_int$P), ]
```

But we don't need these intermediate variables so it's a bit annoying to create them. It also makes the code hard to follow in the next line where we use our intermediate variable.

So what about just overwriting the intermediate variable, like this:
```{r}
sumstatrs_chr1 <- sumstats[sumstats$CHR == 1,]
sumstatrs_chr1 <- sumstatrs_chr1[order(sumstatrs_chr1$P), ]
```

But this isn't really a good idea either, what if you make a mistake you then have to recreate `sumstats_chr1` which could take a while and also your prone to getting in a muddle if you do this (I speak from experience!).

So let's try and get the same output but use pipes instead to make our code much more readable.

```{r}
sumstats_chr1 <- sumstats %>%
                 filter(CHR == 1) %>%
                 arrange(P)
```

This is much easier to read and follow. 

### Problems I'm still trying to figure out the answers for

As you can see when I created the dummy data I could have used pipes then to make it easier to understand how I made `SNP`. But I'm having some issues with this, how can I specify that I want `x` to be the output from the `round(0)` function?

```{r, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE}
x <-1
```

```{r, echo = TRUE, error=FALSE, warning=FALSE, message=FALSE}
SNP <- runif(220, min=11250701, max=185339560) %>%
  round(0) %>%
    paste("rs", x, sep = "") # This line won't work as I haven't defined x.
```

#### Solution 

During coding club session Hon Wah helpfully told me you can use `.` in place of `x` and this will work!

```{r}
SNP <- runif(220, min=11250701, max=185339560) %>%
  round(0) %>%
    paste("rs", ., sep = "") 
SNP[1:10]
```

As I said, only just started playing around with using pipes so would appreciate any tips or reccomendations of `dplyr` functions to use with them to wrangle dataframes in R.

### Resources:

* https://www.datacamp.com/community/tutorials/pipe-r-tutorial
* https://cfss.uchicago.edu/notes/pipes/
* https://uc-r.github.io/pipe
* https://style.tidyverse.org/pipes.html
* https://www.tidyverse.org/packages/



